# Preliminary work on the Chirossis Data set:

```{r}
cir = read.csv("C:\\Users\\lazar\\Downloads\\cirrhosis.csv")

dim(cir)
```


Starting with patients 313 - 418, there is a lack of information on categorical features such as ascites, hepmeg, spiders, and edema. As a result, I will be removing these patients. The reason I am not keeping them is because it will be hard to predict their categorical features, and other mreasurments such as Chol, cu, alkphos, sgot, trig, and plat. There are simply way too many missing patient variables for these

```{r}
cir = cir[-c(313:418),]
```

Now we must convert all of these categorical variables. First, I would like to see what unique categories there are for the categorical variables

```{r}
unique(cir$rx)
unique(cir$sex)
unique(cir$ascites)
unique(cir$hepmeg)
unique(cir$spiders)
unique(cir$edema)

```


Lets one hot encode the variables

```{r}
library(dplyr)

vars_to_encode <- c("rx", "sex", "ascites", "hepmeg", "spiders", "edema")

# Create a new data frame with the one hot encoded variables
one_hot_encoded <- cir %>%
  select(all_of(vars_to_encode)) %>%
  mutate_all(as.factor) %>%
  model.matrix(~.-1, data = .)

# Give meaningful column names to the one hot encoded variables
colnames(one_hot_encoded) <- gsub(".*\\.", "", colnames(one_hot_encoded))

# Combine the original data frame with the one hot encoded variables
cir_encoded <- cbind(cir %>% select(-all_of(vars_to_encode)), one_hot_encoded)

```

We will remove rxplacebo to remove colliniarities
```{r}
cir_encoded <- cir_encoded[, -which(colnames(cir_encoded) == "rxplacebo")]
```



Transforming values from ? to NA
```{r}
cir_encoded$chol <- as.integer(ifelse(cir_encoded$chol == "?", NA, cir_encoded$chol))
cir_encoded$cu <- as.integer(ifelse(cir_encoded$cu == "?", NA, cir_encoded$cu))
cir_encoded$alkphos <- as.integer(ifelse(cir_encoded$alkphos == "?", NA, cir_encoded$alkphos))
cir_encoded$sgot <- as.integer(ifelse(cir_encoded$sgot == "?", NA, cir_encoded$sgot))
cir_encoded$trig <- as.integer(ifelse(cir_encoded$trig == "?", NA, cir_encoded$trig))
cir_encoded$plat <- as.integer(ifelse(cir_encoded$plat == "?", NA, cir_encoded$plat))
cir_encoded$ptt <- as.integer(ifelse(cir_encoded$ptt == "?", NA, cir_encoded$ptt))
cir_encoded$stage <- as.integer(ifelse(cir_encoded$stage == "?", NA, cir_encoded$stage))

```

```{r}
summary(cir_encoded)
```


Usually in SAS, variables listed as ...rxD-peni... mess up the reading of the data, thus I will change it to an underscore

```{r}
names(cir_encoded)[names(cir_encoded) == "rxD-penicillamine"] <- "rxD_penicillamine"
```

event:
censored = 0
liver transplant = 1
dead = 2


We will remove our liver transplant patients.

In addition, I will do the following
censored = 0
dead = 1

```{r}
# Drop value 1
cir_encoded <- cir_encoded[cir_encoded$event != 1,]

# Replace values 0 and 2
cir_encoded$event <- ifelse(cir_encoded$event == 0, 1, 0)

```



Hailey wanted this variable to be added back. 

Basically, if edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics & edemano_edema_and_no_diuretic_therapy_for_edema 0, then the new variable edema_despite_diuretic_therapy will be equal to 1. 

```{r}
library(dplyr)

cir_encoded <- cir_encoded %>%
  mutate(edema_despite_diuretic_therapy = if_else(edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics == 0 & edemano_edema_and_no_diuretic_therapy_for_edema == 0, 1, 0))

```

Now, I will factor the variable. 

```{r}
cir_encoded$rxD_penicillamine = as.factor(cir_encoded$rxD_penicillamine)
cir_encoded$sexmale = as.factor(cir_encoded$sexmale)
cir_encoded$ascitesyes = as.factor(cir_encoded$ascitesyes)
cir_encoded$hepmegyes = as.factor(cir_encoded$hepmegyes)
cir_encoded$spidersyes = as.factor(cir_encoded$spidersyes)
cir_encoded$edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics = as.factor(cir_encoded$edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics)
cir_encoded$edemano_edema_and_no_diuretic_therapy_for_edema = as.factor(cir_encoded$edemano_edema_and_no_diuretic_therapy_for_edema)
```

```{r}
sum(is.na(cir_encoded))
```

We have a total of 62 mising values. Instead of dropping them, we will impute them by using a random forest algorithm. 


I want to see what the distribution looks like before we impute the data

```{r}
library(ggplot2)

# Set the figure size and resolution
options(repr.plot.width=10, repr.plot.height=8, repr.plot.res=300)

# Define the list of excluded variables
excluded_vars <- c("rxD_penicillamine", "sexmale", "ascitesyes", "hepmegyes", "spidersyes", 
                   "edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics", 
                   "edemano_edema_and_no_diuretic_therapy_for_edema")

# Create a list of data frames, each containing one variable and its name, excluding the excluded variables
data_list <- lapply(names(cir_encoded[!(names(cir_encoded) %in% excluded_vars)]), 
                    function(x) data.frame(variable = x, value = cir_encoded[,x]))

# Create a list of ggplot objects, one for each variable
plot_list <- lapply(data_list, function(x) ggplot(x, aes(x = value, fill = variable)) + 
                    geom_density(alpha = 0.5) + 
                    ggtitle(x$variable) +
                    theme(plot.title = element_text(hjust = 0.5)))

# Combine the ggplot objects into a single plot using the grid.arrange function from the gridExtra package
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 3)

```

We will now use the missForest algorithm to imputate the data

```{r}
library(missForest)
imputed_data = missForest(cir_encoded, maxiter = 10)


imp_randomForrest = imputed_data$ximp
```

```{r}
summary(imp_randomForrest)
```

Box plot version of our pdf

```{r}
library(ggplot2)

# Set the figure size and resolution
options(repr.plot.width=10, repr.plot.height=8, repr.plot.res=300)

# Define the list of excluded variables
excluded_vars <- c("rxD_penicillamine", "sexmale", "ascitesyes", "hepmegyes", "spidersyes", 
                   "edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics", 
                   "edemano_edema_and_no_diuretic_therapy_for_edema")

# Create a list of data frames, each containing one variable and its name, excluding the excluded variables
data_list <- lapply(names(imp_randomForrest[!(names(imp_randomForrest) %in% excluded_vars)]), 
                    function(x) data.frame(variable = x, value = imp_randomForrest[,x]))

# Create a list of ggplot objects, one for each variable
plot_list <- lapply(data_list, function(x) ggplot(x, aes(x = variable, y = value)) + 
                    geom_boxplot(fill = "lightblue") + 
                    ggtitle(x$variable) +
                    theme(plot.title = element_text(hjust = 0.5)))

# Combine the ggplot objects into a single plot using the grid.arrange function from the gridExtra package
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 3)

```


pdf of the imputed data 
```{r}
library(ggplot2)

# Set the figure size and resolution
options(repr.plot.width=10, repr.plot.height=8, repr.plot.res=300)

# Define the list of excluded variables
excluded_vars <- c("rxD_penicillamine", "sexmale", "ascitesyes", "hepmegyes", "spidersyes", 
                   "edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics", 
                   "edemano_edema_and_no_diuretic_therapy_for_edema")

# Create a list of data frames, each containing one variable and its name, excluding the excluded variables
data_list <- lapply(names(imp_randomForrest[!(names(imp_randomForrest) %in% excluded_vars)]), 
                    function(x) data.frame(variable = x, value = imp_randomForrest[,x]))

# Create a list of ggplot objects, one for each variable
plot_list <- lapply(data_list, function(x) ggplot(x, aes(x = value, fill = variable)) + 
                    geom_density(alpha = 0.5) + 
                    ggtitle(x$variable) +
                    theme(plot.title = element_text(hjust = 0.5)))

# Combine the ggplot objects into a single plot using the grid.arrange function from the gridExtra package
library(gridExtra)
grid.arrange(grobs = plot_list, ncol = 3)

```


now, I want to superimpose the graphs together to show before and after 
```{r}
library(ggplot2)
library(gridExtra)

# Set the figure size and resolution
options(repr.plot.width=10, repr.plot.height=8, repr.plot.res=300)

# Define the list of excluded variables
excluded_vars <- c("rxD_penicillamine", "sexmale", "ascitesyes", "hepmegyes", "spidersyes", 
                   "edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics", 
                   "edemano_edema_and_no_diuretic_therapy_for_edema")

# Create a list of data frames, each containing one variable and its name, excluding the excluded variables
data_list_1 <- lapply(names(cir_encoded[!(names(cir_encoded) %in% excluded_vars)]), 
                      function(x) data.frame(variable = x, value = cir_encoded[,x], dataset = "original data set"))

data_list_2 <- lapply(names(imp_randomForrest[!(names(imp_randomForrest) %in% excluded_vars)]), 
                      function(x) data.frame(variable = x, value = imp_randomForrest[,x], dataset = "imputed random forest"))

# Combine the data frames
combined_data_list <- mapply(rbind, data_list_1, data_list_2, SIMPLIFY = FALSE)

# Create a list of ggplot objects, one for each variable
plot_list <- lapply(combined_data_list, function(x) ggplot(x, aes(x = value, fill = dataset, color = dataset)) + 
                    geom_density(alpha = 0.5) + 
                    ggtitle(x$variable[1]) +
                    theme(plot.title = element_text(hjust = 0.5)))

# Combine the ggplot objects into a single plot using the grid.arrange function from the gridExtra package
grid.arrange(grobs = plot_list, ncol = 3)

```

looking at placebo vs rxD
```{r}
library(GGally)
imp_randomForrest %>% 
  select(-id) %>% 
  select(age, bili, chol, alb, cu, alkphos, sgot, trig, plat, ptt, stage, rxD_penicillamine, sexmale) %>% 
  ggpairs(aes(fill = rxD_penicillamine))
```

stage graph

```{r}
df <- imp_randomForrest %>% 
  group_by(sexmale, stage) %>% 
  summarise(counts = n())

ggplot(df, aes(x = stage, y = counts)) + 
  geom_bar(aes(fill = sexmale), stat = "identity", position = "dodge") + 
  geom_text(aes(label = counts, group = sexmale), position = position_dodge(0.9), vjust = -.3, size = 3.5) + 
  scale_fill_manual(values = c("#EEAB5F", "#EE5F93"))
```

Survival Analysis 

```{r}
library(survival)
```


```{r}
# Fit the Cox proportional hazards model with stratification by stage
fit <- coxph(Surv(age, event) ~ alb + alkphos + ascitesyes + bili + chol + cu + hepmegyes + plat + ptt + rxD_penicillamine + 
                sexmale + sgot + spidersyes + strata(stage) + trig + edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics + edemano_edema_and_no_diuretic_therapy_for_edema + edema_despite_diuretic_therapy, data = imp_randomForrest)

# Print the model summary
summary(fit)
```

```{r}
# Fit the Cox proportional hazards model with stratification by stage
fit <- coxph(Surv(age, event) ~  cu + hepmegyes + ptt + edemano_edema_and_no_diuretic_therapy_for_edema, data = imp_randomForrest)

# Print the model summary
summary(fit)
```


```{r}
library(survminer)

# Predict the survival probabilities
predicted_survival <- survfit(fit)

# Plot the survival curves
g <- ggsurvplot(
  predicted_survival,
  data = imp_randomForrest,
  pval = TRUE,           # Add p-value
  risk.table = TRUE,     # Add risk table
  conf.int = TRUE,       # Add confidence intervals
  legend.labs = c("Group 1"), # Change legend labels as per your groupings
  palette = c("blue"), # Change colors as desired
  xlab = "Age", # Customize x-axis label
  ylab = "Survival Probability", # Customize y-axis label
  title = "Kaplan-Meier Survival Curve" # Customize the title
)

# Print the plot
g


```

```{r}
# Create separate survival objects for each stage
stage_survival <- survfit(Surv(age, event) ~ stage, data = imp_randomForrest)

# Plot the survival curves
g <- ggsurvplot(
  stage_survival,
  data = imp_randomForrest,
  pval = TRUE,           # Add p-value
  risk.table = TRUE,     # Add risk table
  conf.int = TRUE,       # Add confidence intervals
  legend.labs = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"), # Change legend labels as per your groupings
  palette = c("blue", "red", "green", "purple"), # Change colors as desired
  xlab = "Age", # Customize x-axis label
  ylab = "Survival Probability", # Customize y-axis label
  title = "Kaplan-Meier Survival Curve by Stage" # Customize the title
)

# Print the plot
g


```


```{r}
# Perform the log-rank test
log_rank_test <- survdiff(Surv(age, event) ~ stage, data = imp_randomForrest)

# Print the test results
log_rank_test

```

There are four stages (1, 2, 3, and 4), and their corresponding sample sizes are 16, 64, 112, and 101, respectively.
For each stage, the observed and expected events are listed. For example, in stage 1, there were 15 observed events, whereas 6.3 events were expected under the null hypothesis.
The test statistic contributions ((O-E)^2/E) and the variance contributions ((O-E)^2/V) are also listed for each stage. These values are used to compute the overall chi-squared statistic and its associated p-value.
The overall chi-squared statistic is 40.7 with 3 degrees of freedom. The associated p-value is 8e-09 (which is extremely small).
Since the p-value (8e-09) is much smaller than a typical significance level (e.g., 0.05), you can reject the null hypothesis. This means that there is a statistically significant difference in the survival curves between the different stages. In other words, the survival probabilities are significantly different among the four stages.


```{r}
# Perform pairwise log-rank tests
pairwise_tests <- pairwise_survdiff(Surv(age, event) ~ stage, data = imp_randomForrest, p.adjust.method = "bonferroni")

# Print the results
pairwise_tests

```


Stage 1 vs. Stage 2: The adjusted p-value is 0.38, which is greater than 0.05 (a typical significance level). Therefore, there is no statistically significant difference in the survival curves between Stage 1 and Stage 2.

Stage 1 vs. Stage 3: The adjusted p-value is 0.17, which is also greater than 0.05. This indicates that there is no statistically significant difference in the survival curves between Stage 1 and Stage 3.

Stage 1 vs. Stage 4: The adjusted p-value is 7.3e-08, which is much smaller than 0.05. This means that there is a statistically significant difference in the survival curves between Stage 1 and Stage 4.

Stage 2 vs. Stage 3: The adjusted p-value is 1.00, indicating no statistically significant difference in the survival curves between Stage 2 and Stage 3.

Stage 2 vs. Stage 4: The adjusted p-value is 2.8e-06, which is smaller than 0.05. This implies that there is a statistically significant difference in the survival curves between Stage 2 and Stage 4.

Stage 3 vs. Stage 4: The adjusted p-value is 1.1e-05, which is also smaller than 0.05. This means that there is a statistically significant difference in the survival curves between Stage 3 and Stage 4.

```{r}
# library(ggridges)
# library(ggplot2)
# library(viridis)
# library(hrbrthemes)
# library(reshape2)
# 
# # Select variables to plot
# vars_to_plot <- colnames(imp_randomForrest)[!(colnames(imp_randomForrest) %in% c("rxD-penicillamine", "sexmale", "ascitesyes", "hepmegyes", "spidersyes", "edemaedema_present_without_diuretics_or_edema_resolved_by_diuretics", "edemano_edema_and_no_diuretic_therapy_for_edema"))]
# 
# # Create plot
# ggplot(melt(imp_randomForrest[,vars_to_plot]), aes(x=value, y=variable, fill=value)) + 
#   geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
#   scale_fill_viridis(name = "Density", option = "C") +
#   labs(title = "Density distributions of variables in imp_randomForest") +
#   theme_minimal()

```


```{r}
#write.csv(imp_randomForrest, file ="C:\\Users\\lazar\\Documents\\Spring2023\\Survival_Analysis\\cirrhosisCLEANED.csv" )

```

